---
title: '`shiny`-App Featuring the `swiss` Data Set'
author: "Oliver Stroemer"
date: "2019-07-27"
output: ioslides_presentation
---

## Overview

This `shiny`-app is built to help you explore the `swiss` data set from the `datasets` package via linear regression lines.

You can assign the different variables of the data set to the features of the plot like regressand, regressor, color and size of the geoms.

```{r}
names(swiss)
```

## Example Plot

```{r echo=F}
library(ggplot2)
```

The plots are made using the `ggplot2` package. Here is an example of the plots you can create with this app.

```{r plot, fig.show="hide"}
ggplot(swiss, aes(Agriculture, Fertility)) +
        geom_point(aes(color=Examination, size=Education)) +
        geom_smooth(method="lm")
```

The plot is on the next slide!

***

```{r ref.label="plot", echo=F}
```

Isn't that cool?

## kthxbye

Now you might try out the app; here's the link:

[App on shinyapps.io](https://olistroemer.shinyapps.io/dsc_09-03_project/)

If you are interested in the source code, you can find it here:

[Repository on GitHub](https://github.com/olistroemer/datasciencecoursera/tree/master/09_Developing_Data_Products/Week4_Project)