---
title: "The Effect of Vitamin C on the Tooth Growth of Guinea Pigs"
author: "Oliver Stroemer"
date: "2019-07-16"
output: pdf_document
---

## Synopsis

In this project we are going to explore the ToothGrowth data set from the R
datasets package and compare the tooth growth of guinea pigs in response to the
administration of vitamin C by supplement type and dose via hypothesis tests.

We conclude that the dosage highly affects tooth growth with higher dosage
leading to longer teeth ($p < 0.001$) and that the supplement as orange juice is
more effective than pure ascorbic acid for small dosages ($p = 0.002$).

## Load libraries
```{r}
library(lattice)
```

## Load the data in R

First we need to load the data. There should be 60 observations of 3 variables
as described in the R help page of the ToothGrowth data set obtained by
`?ToothGrowth`.

```{r}
data("ToothGrowth")
dim(ToothGrowth)
```

## Summary of the data

Let's have a quick look at a summary of our data.

```{r}
summary(ToothGrowth)
```

The 60 observations are split into two equally sized groups by the variable
`supp`. Lets see how the length of the teeth relate to the dose grouped by the
supplement type. We want to look at the dose as a factor. Therefore we add
a new column `dosef` to our data before we create our plot.

```{r}
ToothGrowth$dosef <- factor(ToothGrowth$dose)

bwplot(len ~ dosef | supp, data = ToothGrowth, horizontal = F,
       xlab="dose", ylab="length")
```

There is only one outlier in the vitamin C group with 1 mg/day dosage. We can
assume that the length of the teeth follows a normal distribution.

With both types of supplement higher dosage seems to mean longer teeth.

The supplement of vitamin C by orange juice seems to lead to longer teeth than
with ascorbic acid, but only in lower dosages of 0.5 and 1 mg/day dosages; the
difference seems to disappear in a 2 mg/day dosage. Interestingly the variance
for the 2 mg/day seems much higher using ascorbic acid.

## Testing our hypotheses

### More vitamin c means longer teeth

We state that a dose of 2 mg/day of vitamin C of either supplement type leads to
longer teeth. $H_{0}: \mu = \mu_{0}$ vs. $H_{a}: \mu > \mu_{0}$.

An unpaired, one-sided, two sample t-test is used to check our hypothesis.

```{r}
t.test(ToothGrowth[ToothGrowth$dose >  1,]$len,
       ToothGrowth[ToothGrowth$dose <= 1,]$len,
       alternative="greater")
```

With very high confidence $p < 0.001$ we can reject the null hypothesis and
conclude that a dose of 2 mg/day leads to longer teeth than a dose of 0.5
mg/day.

### Supplement by orange juice leads to longer teeth than pure ascorbic acid

We state that vitamin C supplemented by orange juice leads to longer teeth than
pure ascorbic acid if the dose is 0.5 mg/day or 1 mg/day. $H_{0}: \mu = \mu_{0}$
vs. $H_{a}: \mu > \mu_{0}$.

Again an unpaired, one-sided, two sample t-test is used to check our hypothesis.

```{r}
t.test(ToothGrowth[ToothGrowth$dose < 2 & ToothGrowth$supp == "OJ",]$len,
       ToothGrowth[ToothGrowth$dose < 2 & ToothGrowth$supp == "VC",]$len,
       alternative="greater")
```

We can safely reject $H_{0}$ with $p = 0.002$ and conclude that the supplement
by orange juice is more effective than pure ascorbic acid for smaller dosages
under or equal to 1 mg/day.